Fail Faster

### 2025-03-11 paper/paper.tex
msp explanation

This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
\cn{} 
% USE THIS: https://arxiv.org/pdf/2309.08207

\jwc{words}

As an example, consider
\begin{lstlisting}
let is_even x = .< .~x mod 2 == 0 >. in
let succ x = .< 1 + .~x >. in
eta (succ . is_even)
\end{lstlisting}
This is equal to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}: the functions have been fused together, and the overhead of calling them is not present at runtime.

This ``trick'' is how we use metaprogramming

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
    \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value eliminates the function abstraction.
    \end{itemize}
    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
    abstractions the library introduces are fully eliminated at compile time.
  \end{itemize}
}

>>>>

This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
\cn{} 
% USE THIS: https://arxiv.org/pdf/2309.08207
It returns a code for a function that takes an argument \texttt{x}, and then it splats in the result of calling \texttt{f} on just
the quoted \texttt{x}.
\jwc{Does this make any sense?}

The trick is best illustrated by an example. The following program reduces to \texttt{.< fun x -> (1 + x) mod 2 == 0 >}.
\begin{lstlisting}
let is_even x = .< .~x mod 2 == 0 >. in
let succ x = .< 1 + .~x >. in
eta (succ . is_even)
\end{lstlisting}
By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
the functions are fused together... \jwc{words}.
This is the basis of how staging is used to eliminate the abstraction of DSLs
(like a generator DSL). By writing DSL combinators as compile-time functions ---
and only calling \texttt{eta} at the end on the completed DSL program ---  we
can ensure that any overhead of using them is eliminated by run time.

% \jwc{
%   \begin{itemize}
%     \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
%     \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
%     \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
%     \item We can convert one way, but not the other.
%     \item Functions \texttt{'a code -> 'b code} ``fuse''.
%     \begin{itemize}
%       \item Consider writing \texttt{even . succ}. This
%       \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
%       \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
%       a dynamic function value eliminates the function abstraction.
%     \end{itemize}
%     \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
%     abstractions the library introduces are fully eliminated at compile time.
%   \end{itemize}
% }
